---
title: "Finding the earliest pathology in Ataxia"
author: "Roshan Fallah"
date: "2023-05-09"
output:
  prettydoc::html_pretty:
    theme: material
    highlight: github
  shiny::shinyApp:
  html_document:
    toc: true
    toc_float: true   
    animation: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Neurons die in an ataxic pathology 

- Hereditary ataxia is an autosomal neurodegenerative disorder
- Dr De Bellard's lab investigates the pathological mechanisms
- No cure no treatment no isolated or mapped gene
- Identify pathology using Hans Wistar rat models
- Perfusion, cerebellum extraction & IHC
- Determine pathology in 3 rat age groups

```{r}
# Load libraries
# Load first dataset "Total surviving neurons in 3 rat age groups"

library(ggplot2)
library(here)
library(tidyverse)
library(dplyr)
library(gganimate)
```


```{r}
rat_age <- read.csv(here("Data", "rat_agetotalnuclei.csv"))

head(rat_age)
```


```{r echo = FALSE, collapse=TRUE}


plot <- ggplot(rat_age, aes(x = rat_age, y = total.nuclei)) +
  geom_col(aes(fill = factor(rat_age))) +
  scale_fill_manual(values = rainbow(length(unique(rat_age$rat_age)))) +
  scale_y_continuous(limits = c(0, 60))+
  xlab("Rat age group")+
  ylab("Total surviving nuclei")+
  theme(axis.title.y = element_text(color = "maroon", size = 16))+
  theme(axis.title.x = element_text(color = "maroon", size = 16))+
  labs(title = "Total Nuclei by rat age")+
  theme(plot.title = element_text(color = "maroon", size = 18, face = "bold"))+
  theme_dark()


#plot

anim <- plot +
  transition_states(total.nuclei, transition_length = 2, state_length = 1) +
  enter_fade() +
  exit_fade()


animate(anim)  

```


# Hypotheses for determining earliest pathology

- There are pathological changes prior to P15
- Use IHC to count neurons in 5 cerebellar lobes
- loss of purkinje cells (neurons) expected in P10 rats 

```{r}

### Loading 2nd dataset neuronal loss in 5 lobes

lobes <- read.csv(here("Data", "Neuronal_loss_5_lobes.csv"))

#glimpse(lobes)

head(lobes)

lobe<- ggplot(data = lobes, aes(x = Lobe.regions))+
  geom_col(aes(y = WT.total.nuclei, fill = "WT"), position = position_dodge(width = 0.9)) +
  geom_col(aes(y = SP.total.nuclei, fill = "SP"), position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("maroon", "blue"), name = "Spastic & Wildtype") +
  theme(axis.title.y = element_text(color = "maroon", size = 16))+
  theme(axis.title.x = element_text(color = "maroon", size = 16))+
  xlab("5 lobe regions tested") +
  ylab("Total Nuclei found") +
  ggtitle("Loss of neurons in p15 ataxic rats")+
  theme(plot.title = element_text(color = "maroon", size = 18, face = "bold"))


lobe

```


# Measuring soma size for mutant P10 vs wildtype

- Using IHC and fluorescent microscopy to measure purkinje cells
- Looking for a reduction in soma size in mutant P10 HW rats


```{r}

# Loading 3rd dataset for P10 rat soma size

soma_size <- read.csv(here("Data", "p10_soma_size.csv"))


soma_rounded <- round(soma_size, digits = 0)  ## rounding all values in both columns

#soma_rounded

head(soma_rounded)

somaround <- soma_rounded %>% rownames_to_column(var = "row_num") # using dplyr to add column with row numbers

somaplot <- ggplot(somaround, aes(x = row_num, y = Wildtype.P10)) + 
  geom_point(aes(color = "Wildtype"), size = 3) + 
  geom_point(aes(y = Spastic.P10, color = "Spastic"), shape = 15, fill = "red", size = 4) +
  scale_color_manual(values = c("Wildtype" = "blue", "Spastic" = "red")) +
  labs(x = "observation no. of measured neurons", y = "Diameter of soma (μm)", color = "red")+
  theme_gray()+
  theme(axis.title.y = element_text(color = "orange", size = 16))+
  theme(axis.title.x = element_text(color = "orange", size = 16))+
  scale_x_discrete(breaks = seq(0, nrow(somaround), by = 18))+
  ggtitle("Scatterplot of soma size in mutant p10 rats")+
  theme(plot.title = element_text(color = "orange", size = 20, face = "bold"))
somaplot


animsoma <- ggplot(somaround, aes(x = row_num, y = Wildtype.P10)) + 
  geom_point(aes(color = "Wildtype"), size = 5) + 
  geom_point(aes(y = Spastic.P10, color = "Spastic"), shape = 17, fill = "red", size = 7) +
  scale_color_manual(values = c("Wildtype" = "blue", "Spastic" = "red")) +
  labs(x = "observation no. of measured neurons", y = "Diameter of soma (μm)", color = "red") +
  theme_gray()+
  theme(axis.title.y = element_text(color = "orange", size = 16))+
  theme(axis.title.x = element_text(color = "orange", size = 16))+
  scale_x_discrete(breaks = seq(0, nrow(somaround), by = 20))+
  ggtitle("Scatterplot of soma size in mutant p10 rats")+
  theme(plot.title = element_text(color = "orange", size = 20, face = "bold"))
  

animated <- animsoma +
  transition_time(Spastic.P10) +
  ease_aes('linear')
animated


```

# Discussion

- hypotheses supported pathology seen in P10 rats
- Further testing on P10 to reinforce findings
- Future experiments on <P7